The Name of the Title Is Hope

### 2025-01-22 paper.tex
Revisions

\usepackage[ruled,vlined]{algorithm2e}
\usepackage{amsmath}
\usepackage{mathtools}
\newcommand{\thomas}[1]{{\footnotesize\color{orange}[Thomas: #1]}}
\newcommand{\john}[1]{{\footnotesize\color{cyan}[John: #1]}}
\newcommand{\raph}[1]{{\footnotesize\color{magenta}[Raph: #1]}}

@@

\subsubsection{Inter-Workgroup Barriers}
Although threads \emph{within} a workgroup may synchronize with each other by participating in a \emph{barrier} primitive, the virtualization of processors introduces an emergent high-level limitation: the absence of an inter-workgroup barrier. As previously mentioned, workgroups cannot context switch. Although contemporary GPU hardware~\cite{} supports workgroup preemption for prioritizing real-time tasks or managing multi-process workloads, this functionality is deliberately excluded from the GPU programming model. This stems from the prohibitively high overhead in latency and storage associated with moving workgroup contexts on and off chip. In the oversubscription paradigm, where the number of workgroups is theoretically unbound and typically far exceeds the available multiprocessors, every invocation of such a barrier would necessitate context-switching between all launched workgroups. Due to this high cost, no GPU programming environment furnishes developers with a true unbounded-launch inter-workgroup barrier.\footnote{CUDA recently introduced \emph{Thread Block Cluster} synchronization~\cite{}, but it is almost certainly designed to operate within the \emph{Persistent-Thread}~\cite{} paradigm, rather than serving as a barrier across an unbounded launch.} Instead, when inter-workgroup synchronization is necessary, developers must rely on kernel launches as synchronization points. However, kernel launches are suboptimal as barriers: they incur overheads due to the heterogeneous nature of the operation, and all execution context is lost between launches.

\subsubsection{Fairness and Progress}
At the subgroup scheduling level, two key issues emerge: \emph{fairness}---how evenly execution resources are distributed among threads---and \emph{progress guarantees}---ensuring that all threads or subgroups eventually make progress towards termination. However, the term "fairness" is used differently across the literature, leading to potential confusion. Sorensen et al.~\cite{10.1145/3022671.2984032, sorensen_et_al:LIPIcs.CONCUR.2018.23, 10.1145/3485508}, whose work provides the most comprehensive taxonomy of progress models to date, use "fairness" specifically to describe differing levels of progress guarantees. In contrast, NVIDIA publications~\cite{4523358, Merrill2016} define "fairness" in terms of the distribution of execution resources among subgroups. In this work, we adopt NVIDIA's terminology to maintain consistency with their publications, while acknowledging the broader definitions provided by Sorensen et al.

@@

\subsubsection{Scan-then-Propagate}
Introduced by Sengupta et al.~\cite{10.5555/1280094.1280110}, the \emph{Scan-then-Propagate}~\cite{GPUGems3, Sengupta2008} strategy is an intuitive extension of the intra-workgroup scan, comprising three phases:
\begin{enumerate}
  \item \textbf{Intra-Workgroup (Local) Scan:} Each workgroup performs an inclusive scan on its assigned work tile partition of the input data. The results, including both the local scan output and the reduction of each workgroup, are written back to global memory.
  \item \textbf{Spine Scan:} The reductions from all workgroups, collectively referred to as \emph{spine}, are gathered and scanned to compute the global prefix of the reductions, resolving the dependencies between work tiles; we call the result of this phase the \emph{root}. For large scans, this stage is an insignificant amount of work compared to the first and third.
  \item \textbf{Propagation:} Each workgroup retrieves its portion of the root and combines it with the result of the local scan to compute the final scan.
\end{enumerate}

@@

\subsubsection{Reduce-then-Scan}
If memory bandwidth resources are sufficiently scarce relative to compute, memory-bound algorithms like scan can profit by trading additional computation for reduced memory traffic. This trade-off led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{Merrill-Grimshaw, Ha-and-Han, Dotsenko, Breitbart}: by writing only the work tile reduction to global memory in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated. Like \emph{Scan-then-Propagate}, \emph{Reduce-then-Scan} employs a three-phase approach using kernel launches to resolve dependencies.
\begin{enumerate}
  \item \textbf{Work Tile Reduction:} Each workgroup performs a pure reduction on its work tile partition, posting the result to global memory.
  \item \textbf{Spine Scan:} The spine is gathered and scanned to produce the root.
  \item \textbf{Intra-Workgroup Scan and Propagation:} Each workgroup performs the local scan on the work tile, incorporating its portion of the root as it writes to global memory.
\end{enumerate}

Merill and Grimshaw~\cite{} introduced a workgroup \emph{raking} technique that elides the overhead incurred when the recursive depth exceeds two. In this approach, the total workgroup occupancy $o$ is determined, and the input is partitioned into non overlapping blocks of size $\frac{n}{t \cdot o}$, which are raked serially by each workgroup. This limits the size of the spine from variably sized, proportional to the input size (possibly requiring recursion), to constant sized, and guaranteed to avoid further recursion. Assuming $t \geq o$, global memory movement during the spine scan is reduced from $O(n/t)$ to $O(2o)$. Kernel launches are limited to three and total global memory movement is reduced to $O(3n + 4o) = O(3n)$.

Breitbart~\cite{} was the first to apply more advanced inter-workgroup synchronization techniques to scan. Recognizing that workgroup raking naturally aligns with a \emph{Persistent-Threads}~\cite{} paradigm, Breitbart utilized an atomics-based inter-workgroup barrier to replace kernel launches. Since multiprocessors are not oversubscribed, workgroups remain resident (persist) on the multiprocessors for the entire lifetime of the kernel, enabling synchronization at the barrier without requiring context switching. Although global data movement remains at $O(3n)$, this consolidates the scan to a single kernel, minimizing launch overheads. \thomas{This also requires FPG, but not ideal to unpack and mention here . . .}

\subsubsection{Chained-Scan, Stream-Scan}
First introduced by Yan et al.~\cite{} in \emph{Stream-Scan}, the key innovation of the \emph{Chained-Scan} architecture lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level. Instead of using kernel launches as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups, utilizing atomics and bit-packed status flags to ensure coherent views of data. To enforce this serial ordering, work tiles are dynamically assigned using atomic increment operations, rather than arbitrary assignment based on virtualized workgroup indices. This shifts dependency resolution from explicit synchronization via barrier-like structures to implicit synchronization via the order in which workgroups access global data. As a result, each element is read and written exactly once, achieving the theoretical minimum of $2n$.

@@

While \emph{StreamScan} successfully reduces per-element device memory accesses to two, it does not consistently saturate global memory bandwidth, preventing it from achieving speed-of-light performance. This limitation arises because a workgroup in \emph{StreamScan} cannot post its reduction to global memory until it has received the reduction of its predecessor. While this results in efficient $O(n/t)$ communication of reductions, and $O(2n+ n/t)= O(2n)$ total data movement, the sequential dependency creates a bottleneck that limits performance to the rate at which reductions propagate through device memory. Although this can be mitigated by increasing the size of the work tile, limited on-chip resources preclude scaling the tile size indefinitely. Moreover, this approach is unsuitable for a portable environment, as excessively large work tiles risk reducing occupancy on less capable hardware.

\subsubsection{Chained-Scan, Decoupled-Lookback}
Merrill and Garland~\cite{} introduced the \emph{Decoupled-Lookback} technique, which separates reduction propagation and dependency resolution to address the bottleneck caused by propagation latency. In this approach, each workgroup posts its local reduction to global memory and then performs a \emph{lookback}---a serial traversal backward along the spine---to compute its inclusive reduction. Work and memory accesses are constant-bounded to the workgroup occupancy $o$ by posting the inclusive reduction after the lookback. However, granting each workgroup control over its dependency resolution does not eliminate reliance on \emph{progress guarantees}, as workgroups must still spin on tiles that have yet to post their reductions, risking deadlock.

@@

\subsection{Ideal Scan Architecture}
\thomas{This will be rewritten. It is clear to me now that the focus should be on the tradeoffs between the Oversubcription and not-Oversubcription(Persistent-Thread or Fixed Launch) paradigms. We want to communicate to the reader that because we want portability, we want to remain within the Oversubcription paradigm. We LIKE Oversubcription. We want to reframe the downsides as "things that happen when you step outside of Oversubcription." I.e. when you step outside of Oversubcription, now YOU have to manage occupancy -> occupancy discovery. When you step outside of Oversubcription now YOU have to worry about launch-related global memory access paterns -> overpartitioning your rake. Then we want to say although Oversubcription precludes inter-workgroup barriers besides kernel launches, thats OK because Chained Scan does not require an explicit barrier. Thus, Chained-Scan is good \emph{because} it lets us stay Oversubscribed}.

In addition to its speed, the \emph{Chained-Scan} architecture has more practical properties that improve over its predecessors. As it is a single kernel, it minimizes launch overheads and simplifies maintenance. On previous architectures, handling of arbitrary input sizes can be awkward: the recursive approach, for example, entails multiple kernel launches and a heterogeneous system to coordinate them. Similarly, the raking approach—a necessary optimization in methods prior to \emph{Chained-Scan}—can be cumbersome, as predetermining static workgroup occupancies for every target device is infeasible. Previous workgroup occupancy discovery techniques, such as the one proposed by Sorensen et al.~\cite{}, are impossible to implement in WebGPU as they require at least a memory fence, which is not supported by the specification.\footnote{Our implementation of \emph{Reduce-Then-Scan}, which can be found in the artifact, includes a fenceless occupancy discovery method amenable to the WebGPU environment, but its description is beyond the scope of this work.}

Additionally, on GPUs where the global memory data cache is indexed by physical addresses—and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)—raking the workgroups can lead to performance degradation due to cache thrashing. This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups. Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns---or applications like radix sorting, where the scattering phase of each pass results in $radixDigits \cdot o$ unique memory locations---may experience significant performance degradation.

As a consequence of its single kernel execution and inherently serial ordering, \emph{Chained-Scan} is also suitable for \emph{in-situ} compaction and allocation. In previous architectures, such tasks required additional kernel launches and bookkeeping to handle dependencies between workgroups. By contrast, \emph{Chained-Scan} efficiently propagates dependencies within a single kernel, enabling its application to a variety of practical routines, including data deduplication, sparse data compaction, error correction, and interval merging.

But, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in terms of its theoretical and implementation characteristics, it falls short in meeting the criterion of \emph{portability}. This limitation arises because \emph{Chained-Scan} depends on a \emph{forward-progress guarantee}, a feature provided by NVIDIA devices but notably lacking on Apple's M-series and ARM GPUs~\cite{sorensen2}. Without this guarantee, \emph{Chained-Scan} is highly prone to catastrophic failure in the form of deadlocking.

\subsection{Why does \emph{Chained-Scan} Rely on Forward-Progress Guarantees?}
The challenges faced by \emph{Chained-Scan} are analogous to the \emph{Dining Philosophers Problem}, where tasks compete for shared resources, potentially resulting in deadlock. Coffman et al.~\cite{10.1145/356586.356588} define four conditions necessary for deadlock:
\begin{enumerate}
  \item \textbf{Mutual Exclusion}: Tasks claim exclusive control of resources they require.
  \item \textbf{Resource Holding}: Tasks hold acquired resources while waiting.
  \item \textbf{No Preemption}: A resource cannot be removed from a task before completion.
  \item \textbf{Circular Dependency}:
        \begin{enumerate}
          \item Dependency Relationships: A dependency exists such that each task holds one or more resources that are required by proceeding tasks.
          \item Circular Ordering: A circular chain of tasks exists.
        \end{enumerate}
\end{enumerate}
Applying this schema to \emph{Chained-Scan}, workgroups correspond to "tasks" and work-tile reductions to "resources." Conditions 1–3 are satisfied due to the one-to-one mapping of work tiles to workgroups and the constraints of the GPU programming model. Similarly, condition 4a is met by the dependency relationships between work tiles, leaving condition 4b---circular ordering---as the key concern.

Although the serial assignment of work tiles ensures that predecessor work tiles are either scheduled or completed, there is no guarantee that the corresponding reductions will be available before a workgroup begins its lookback. This uncertainty arises because workgroups are scheduled individually, and a predecessor workgroup might not yet have had the opportunity to compute and post its reduction. In such cases, the dependent workgroup enters a waiting state, spinning until the reduction becomes available. During this waiting period, the spinning workgroup remains scheduled, consuming computational resources. Without \emph{forward-progress guarantees}, the scheduler may indefinitely prioritize the spinning workgroup over its predecessor, preventing the predecessor from completing its reduction. This creates a circular dependency, satisfying condition 4b and resulting in deadlock within the kernel.

A \emph{forward-progress guarantee} (FPG) is a promise provided by the subgroup scheduler that all occupant subgroups (of possibly different workgroups) eventually make progress toward termination, preventing scenarios where a subgroup is starved of execution time. Sorensen et al.~\cite{barriers, 1, 2} provide the most comprehensive analysis of GPU schedulers. However, while they classify \emph{Chained-Scan} as an FPG-reliant algorithm, their primary focus is in exploring and formalizing progress models, rather than their application to any one algorithm. In the context of \emph{Chained-Scan} deadlocking, FPG prevents condition 4b by prohibiting indefinite scheduling of any single subgroup, thereby breaking the circular dependency chain.
\thomas{Sorensen places Chained-Scan into the HSA category, but it is unclear if this is correct. The definition of HSA is \newline "Threads for which fair execution is guaranteed: The thread with the lowest id that has not yet terminated" \newline However, as Chained-Scan and other serially ordered algorithms no longer use the virtual ID provided by the workgroup scheduler, is fair execution still guaranteed under HSA? It seems that it should be classified under OBE\@.}

\thomas{This example feels redundant now. Timeout detection and recovery can be moved into the 2nd paragraph in this section.}
For example, consider a scan operation requiring two workgroups running on a GPU with a single multiprocessor. Although the multiprocessor can host both workgroups, it has only one scheduling unit, which lacks FPG\@. In this setup, both workgroups launch and acquire their respective work tiles. The workgroup responsible for \emph{Work Tile 1} finishes first and begins spinning, loading the value at the memory address for the reduction of \emph{Work Tile 0} with each iteration. However, because the scheduler lacks FPG, it may continue scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction. As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery} (TDR) and crashing the host program.

@@

\section{Decoupled-Fallback}
\thomas{I've broken somewhat with the outline and moved the goals and non-goals section to somewhere in the introduction. I kept finding myself trying to write around it. I think the content there is good, but it's very awkward as it ended up so deep within the paper. The placement of this section here feels more natural, and seems to mirror other papers?}
\emph{Decoupled-Fallback} enhances the \emph{Scalar-Fallback} technique with significantly improved fallback performance. As an extension of the \emph{Chained-Scan} architecture, it preserves the advantages in speed and oversubscription, making it an ideal choice for our requirements.

\thomas{What I want to accomplish here is to go beyond saying "Here is our method. It works great." I want to show that these are the possible avenues to attack deadlocking . . . and exhaustively demonstrate that removing mutual exclusion is the best way to do it, under our given constraints.}
Like \emph{Scalar-Fallback}, we choose \emph{mutual exclusion} as our approach to break deadlocking. \emph{Dependency relationships} and \emph{no preemption} are immutable characteristics of scan and the GPU programming model, eliminating them as viable avenues for breaking deadlocking. While it may be possible to break \emph{circular ordering} by designing tile sizes such that workgroups are mapped one-to-one to multiprocessors, doing so portably is infeasible. Moreover, vendor guidance on progress guarantees among multiprocessors is also lacking, leaving the same risks of deadlocking but at the multiprocessor level. While eliminating \emph{resource holding}---allowing workgroups to discard work tiles when they encounter a deadlock---is theoretically possible, it is strictly worse than eliminating \emph{mutual exclusion}, as it necessitates $t$ additional movements (discard-read-read versus read).
\begin{algorithm}[htb]
  \small
  \SetAlgoLined
  \KwIn{%
    \texttt{in}[], \texttt{out}[],
    \texttt{tileBump},
    \texttt{spine}[],
    \texttt{wg\_fallback}[],
    \texttt{wg\_lock},
    \texttt{wg\_partials}[],
    \texttt{wg\_prev\_reduction},
    \texttt{wg\_broadcast},
  }
  \KwOut{\texttt{out}[] fully inclusive scan of \texttt{in}[]}

  \If{$\textnormal{threadid.x} = 0$}{
    \texttt{wg\_broadcast} $\gets$ \texttt{atomicAdd}(\&\texttt{tileBump}, 1)\;
    \texttt{wg\_lock} $\gets$ \textnormal{LOCKED}\;
  }
  \textbf{barrier()}\;

  \texttt{part\_id} $\gets$ \texttt{wg\_broadcast}\;

@@

  \textbf{Load}(\texttt{in}, \texttt{part\_id}, \textit{local\_vectors})\;
  \textbf{SubgroupRake}(\textit{local\_vectors})\;

  \If{$\text{\normalfont highestRankLane}(\textnormal{threadid.x})$}{
    \texttt{wg\_partials}[\textnormal{subgroup\_id}] $\gets$ \textit{local\_vectors}[\textnormal{final\_partial}]\;
  }
  \textbf{barrier()}\;

  \textbf{WorkgroupWideScan}(\texttt{wg\_partials})\;
  \textbf{barrier()}\;

>>>>

\usepackage[ruled,vlined]{algorithm2e}
\SetKwInput{KwGlobalIn}{Global Mem Input}
\SetKwInput{KwSharedIn}{Shared Mem Input}
\SetKwInput{KwSharedOut}{Shared Mem Output}
\SetKwInput{KwGlobalOut}{Global Mem Output}
\SetKwInput{KwConstants}{Constants}

\usepackage{amsmath}
\usepackage{mathtools}
\newcommand{\thomas}[1]{{\footnotesize\color{orange}[Thomas: #1]}}
\newcommand{\john}[1]{{\footnotesize\color{cyan}[John: #1]}}
\newcommand{\raph}[1]{{\footnotesize\color{magenta}[Raph: #1]}}

@@

\subsubsection{Inter-Workgroup Barriers}
Although threads \emph{within} a workgroup may synchronize with each other by participating in a \emph{barrier} primitive, the virtualization of processors introduces an emergent high-level limitation: the absence of an inter-workgroup barrier. As previously mentioned, workgroups cannot context switch. Although contemporary GPU hardware~\cite{} supports workgroup preemption for prioritizing real-time tasks or managing multi-process workloads, this functionality is deliberately excluded from the GPU programming model. This stems from the prohibitively high overhead in latency and storage associated with moving workgroup contexts on and off chip. In the oversubscription paradigm, where the number of workgroups is theoretically unbound and typically far exceeds the available multiprocessors, every invocation of such a barrier would necessitate context-switching between all launched workgroups. Due to this high cost, no GPU programming environment furnishes developers with a true unbounded-launch inter-workgroup barrier.\footnote{CUDA recently introduced \emph{Thread Block Cluster} synchronization~\cite{NvidiaCudaGuide}, but it is almost certainly designed to operate within the \emph{Persistent-Thread}~\cite{gupta2012} paradigm, rather than serving as a barrier across an unbounded launch.} Instead, when inter-workgroup synchronization is necessary, developers must rely on kernel launches as synchronization points. However, kernel launches are suboptimal as barriers: they incur overheads due to the heterogeneous nature of the operation, and all execution context is lost between launches.

\subsubsection{Fairness and Progress}
Because multiple workgroups may reside on a single multiprocessor, two key issues arise at the subgroup scheduling level: \emph{fairness}---how evenly execution resources are distributed among threads---and \emph{progress guarantees}---ensuring that all threads or subgroups eventually make progress towards termination. However, the term "fairness" is used differently across the literature, leading to potential confusion. Sorensen et al.\cite{sorensen2016,sorensen2018,sorensen2021}, whose work provides the most comprehensive taxonomy of progress models to date, use "fairness" specifically to describe differing levels of progress guarantees. In contrast, NVIDIA publications~\cite{4523358,Merrill2016} define "fairness" in terms of the distribution of execution resources among subgroups. In this work, we adopt NVIDIA's terminology to maintain consistency with their publications, while acknowledging the broader definitions provided by Sorensen et al.

@@

\subsubsection{Scan-then-Propagate}
Introduced by Sengupta et al.~\cite{10.5555/1280094.1280110}, the \emph{Scan-then-Propagate}~\cite{GPUGems3, Sengupta2011} strategy is an intuitive extension of the intra-workgroup scan, comprising three phases:
\begin{enumerate}
  \item \textbf{Intra-Workgroup (Local) Scan:} Each workgroup performs an inclusive scan on its assigned work tile partition of the input data. The results, including both the local scan output and the reduction of each workgroup, are written back to global memory.
  \item \textbf{Spine Scan:} The reductions from all workgroups, collectively referred to as \emph{spine}, are gathered and scanned to compute the global prefix of the reductions, resolving the dependencies between work tiles; we call the result of this phase the \emph{root}. For large scans, this stage is an insignificant amount of work compared to the first and third.
  \item \textbf{Propagation:} Each workgroup retrieves its portion of the root and combines it with the result of the local scan to compute the final scan.
\end{enumerate}

@@

\subsubsection{Reduce-then-Scan}
If memory bandwidth resources are sufficiently scarce relative to compute, memory-bound algorithms like scan can profit by trading additional computation for reduced memory traffic. This trade-off led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{10.1145/1375527.1375559, Merrill2009, 10.1109/TPDS.2012.336, 10.5555/2031978.2032029}: by writing only the work tile reduction to global memory in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated. Like \emph{Scan-then-Propagate}, \emph{Reduce-then-Scan} employs a three-phase approach using kernel launches to resolve dependencies.
\begin{enumerate}
  \item \textbf{Work Tile Reduction:} Each workgroup performs a pure reduction on its work tile partition, posting the result to global memory.
  \item \textbf{Spine Scan:} The spine is gathered and scanned to produce the root.
  \item \textbf{Intra-Workgroup Scan and Propagation:} Each workgroup performs the local scan on the work tile, incorporating its portion of the root as it writes to global memory.
\end{enumerate}

Merill and Grimshaw~\cite{Merrill2009} introduced a workgroup \emph{raking} technique that elides the overhead incurred when the recursive depth exceeds two. In this approach, the total workgroup occupancy $o$ is determined, and the input is partitioned into non overlapping blocks of size $\frac{n}{t \cdot o}$, which are raked serially by each workgroup. This limits the size of the spine from variably sized, proportional to the input size (possibly requiring recursion), to constant sized, and guaranteed to avoid further recursion. Assuming $t \geq o$, global memory movement during the spine scan is reduced from $O(n/t)$ to $O(2o)$. Kernel launches are limited to three and total global memory movement is reduced to $O(3n + 4o) = O(3n)$.

Breitbart~\cite{10.5555/2031978.2032029} was the first to apply more advanced inter-workgroup synchronization techniques to scan. Recognizing that workgroup raking naturally aligns with a \emph{Persistent-Threads}~\cite{gupta2012} paradigm, Breitbart utilized an atomics-based inter-workgroup barrier to replace kernel launches. Since multiprocessors are not oversubscribed in raking, workgroups remain resident (persist) on the multiprocessors for the entire lifetime of the kernel, enabling synchronization at the barrier without requiring context switching. Although global data movement remains at $O(3n)$, this consolidates the scan to a single kernel, minimizing launch overheads. \thomas{This also requires FPG, but not ideal to unpack and mention here . . .}

\subsubsection{Chained-Scan, Stream-Scan}
First introduced by Yan et al.~\cite{10.1145/2442516.2442539} in \emph{Stream-Scan}, the key innovation of the \emph{Chained-Scan} architecture lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level. Instead of using kernel launches as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups, utilizing atomics and bit-packed status flags to ensure coherent views of data. To enforce this serial ordering, work tiles are dynamically assigned using atomic increment operations, rather than arbitrary assignment based on virtualized workgroup indices. This shifts dependency resolution from explicit synchronization via barrier-like structures to implicit synchronization via the order in which workgroups access global data. As a result, each element is read and written exactly once, achieving the theoretical minimum of $2n$.

@@

While \emph{StreamScan} successfully reduces per-element device memory accesses to two, it does not consistently saturate global memory bandwidth, preventing it from achieving speed-of-light performance. This limitation arises because a workgroup in \emph{StreamScan} cannot post its reduction to global memory until it has received the result of its immediate predecessor. While this results in efficient $O(n/t)$ communication of reductions, and $O(2n+ n/t)= O(2n)$ total data movement, the sequential dependency creates a bottleneck that limits performance to the rate at which reductions propagate through device memory. Although this can be mitigated by increasing the size of the work tile, limited on-chip resources preclude scaling the tile size indefinitely. Moreover, this approach is unsuitable for a portable environment, as excessively large work tiles risk reducing occupancy on less capable hardware.

\subsubsection{Chained-Scan, Decoupled-Lookback}
Merrill and Garland~\cite{Merrill2016} introduced the \emph{Decoupled-Lookback} technique, which separates reduction propagation and dependency resolution to address the bottleneck caused by propagation latency. In this approach, each workgroup posts its local reduction to global memory and then performs a \emph{lookback}---a serial traversal backward along the spine---to compute its inclusive reduction. Work and memory accesses are constant-bounded to the workgroup occupancy $o$ by posting the inclusive reduction after the lookback. However, granting each workgroup control over its dependency resolution does not eliminate reliance on \emph{progress guarantees}, as workgroups must still spin on tiles that have yet to post their reductions, risking deadlock.

@@

\subsection{Ideal Scan Architecture}
The design of an ideal scan architecture must balance performance, portability, and practicality. Beyond their differences in speed, \emph{Chained-Scan} and \emph{Reduce-Then-Scan} embody two fundamentally contrasting paradigms. By eliding explicit inter-workgroup synchronization, \emph{Chained-Scan} aligns naturally with the oversubscription paradigm, delegating occupancy management to the hardware scheduler. In contrast, the raking approach inherent to \emph{Reduce-Then-Scan} mandates the fixed-launch paradigm, necessitating explicit control over occupancy and potentially synchronization.

Fixed-launch approaches demand substantial developer effort, requiring either tuning of workgroup launch sizes across target devices or occupancy discovery. Moreover, previous workgroup occupancy discovery techniques, such as the one proposed by Sorensen et al.~\cite{sorensen2016}, are infeasible to implement in WebGPU due to their reliance on memory fences---a feature not supported by the specification.\footnote{Our implementation of \emph{Reduce-Then-Scan}, which can be found in the artifact, includes a fenceless occupancy discovery method amenable to the WebGPU environment, but its description is beyond the scope of this work.}

Hardware-level considerations further underscore the limitations of fixed-launches. On GPUs where the global memory data cache is indexed by physical addresses---and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)---fixed-launches can lead to performance degradation due to cache thrashing. This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups. Although this behavior typically does not manifest with simple monoids, more complex monoids with \emph{struct-of-array} memory access patterns---or applications like radix sorting, where the scattering phase of each pass results in $radixDigits \cdot o$ unique memory locations---may experience significant performance degradation.

Thus, oversubcription and \emph{Chained-Scan} emerges as the preferred architecture, as it naturally aligns with portability by leveraging hardware schedulers to dynamically manage occupancy across diverse architectures without requiring developer intervention. Furthermore, as a single kernel with inherently serial ordering, \emph{Chained-Scan} is suitable for \emph{in-situ} operations, augmenting practical routines such as data deduplication, sparse data compaction, error correction, and interval merging.

But, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in terms of its theoretical and implementation characteristics, it falls short in fully meeting the criterion of \emph{portability}. This limitation arises because \emph{Chained-Scan} depends on a \emph{forward-progress guarantee}, a feature provided by NVIDIA devices but notably lacking on Apple's M-series and ARM GPUs~\cite{sorensen2021}. Without this guarantee, \emph{Chained-Scan} is highly prone to catastrophic failure in the form of deadlocking.

\subsection{Why does \emph{Chained-Scan} Rely on Forward-Progress Guarantees?}
A \emph{forward-progress guarantee} (FPG) is a promise provided by the subgroup scheduler that all occupant subgroups (of possibly different workgroups) eventually make progress toward termination, preventing scenarios where a subgroup is starved of execution time. Sorensen et al.~\cite{sorensen2018,sorensen2021} provide the most comprehensive analysis of GPU schedulers. However, while they classify \emph{Chained-Scan} as an FPG-reliant algorithm, their primary focus is in exploring and formalizing progress models, rather than their application to any one algorithm.
\thomas{Sorensen places Chained-Scan into the HSA category, but it is unclear if this is correct. The definition of HSA is \newline "Threads for which fair execution is guaranteed: The thread with the lowest id that has not yet terminated" \newline However, as Chained-Scan and other serially ordered algorithms no longer use the virtual ID provided by the workgroup scheduler, is fair execution still guaranteed under HSA? It seems that it should be classified under OBE\@.}

The challenges faced by \emph{Chained-Scan} are analogous to the \emph{Dining Philosophers Problem}, where tasks compete for shared resources, potentially resulting in deadlock. Coffman et al.~\cite{10.1145/356586.356588} define four conditions necessary for deadlock:
\begin{enumerate}
  \item \textbf{Mutual Exclusion}: Tasks claim exclusive control of resources they require.
  \item \textbf{Resource Holding}: Tasks hold acquired resources while waiting.
  \item \textbf{No Preemption}: A resource cannot be removed from a task before completion.
  \item \textbf{Circular Dependency}:
        \begin{enumerate}
          \item Dependency Relationships: A dependency exists such that each task holds one or more resources that are required by proceeding tasks.
          \item Circular Ordering: A circular chain of tasks exists.
        \end{enumerate}
\end{enumerate}
Applying this schema to \emph{Chained-Scan}, workgroups correspond to "tasks" and work-tile reductions to "resources." Conditions 1–3 are satisfied due to the one-to-one mapping of work tiles to workgroups and the constraints of the GPU programming model. Similarly, condition 4a is met by the dependency relationships between work tiles, leaving condition 4b---\emph{circular ordering}---as the key concern.

Although the serial assignment of work tiles ensures that predecessor work tiles are either scheduled or completed, there is no guarantee that the corresponding reductions will be available before a workgroup begins its lookback. This uncertainty arises because workgroups are scheduled individually, and a predecessor workgroup might not yet have had the opportunity to compute and post its reduction. In such cases, the dependent workgroup enters a waiting state, spinning until the reduction becomes available. During this waiting period, the spinning workgroup remains scheduled, consuming computational resources. Without FPG, the scheduler may indefinitely prioritize the spinning workgroup over its predecessor, preventing the predecessor from completing its reduction. This creates a circular dependency, satisfying condition 4b and resulting in deadlock within the kernel.

For example, consider a scan operation requiring two workgroups running on a GPU with a single multiprocessor. Although the multiprocessor can host both workgroups, it has only one scheduling unit, which lacks FPG\@. In this setup, both workgroups launch and acquire their respective work tiles. The workgroup responsible for \emph{Work Tile 1} finishes first and begins spinning, loading the value at the memory address for the reduction of \emph{Work Tile 0} with each iteration. However, because the scheduler lacks FPG, it may continue scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction. As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery} (TDR) and crashing the host program.

@@

\section{Decoupled-Fallback}
\thomas{I've broken somewhat with the outline and moved the goals and non-goals section to somewhere in the introduction. I kept finding myself trying to write around it. I think the content there is good, but it's very awkward as it ended up so deep within the paper. The placement of this section here feels more natural, and seems to mirror other papers?}
\emph{Decoupled-Fallback} separates fallback operations from the lookback process, enhancing the \emph{Scalar-Fallback} technique with significantly improved fallback performance. As an extension of the \emph{Chained-Scan} architecture, it retains the advantages of high speed and oversubscription, aligning with our design requirements.

\thomas{What I want to accomplish here is to go beyond saying "Here is our method. It works great." I want to show that these are the possible avenues to attack deadlocking . . . and exhaustively demonstrate that removing mutual exclusion is the best way to do it, under our given constraints.}
Like \emph{Scalar-Fallback}, we choose \emph{mutual exclusion} as our approach to break deadlocking. \emph{Dependency relationships} and \emph{no preemption} are immutable characteristics of scan and the GPU programming model, eliminating them as viable avenues for breaking deadlocking. While it may be possible to break \emph{circular ordering} by designing tile sizes such that workgroups are mapped one-to-one to multiprocessors, doing so portably is infeasible. Moreover, vendor guidance on progress guarantees among multiprocessors is also lacking, leaving the same risks of deadlocking but at the multiprocessor level. While eliminating \emph{resource holding}---allowing workgroups to discard work tiles when they encounter a deadlock---is theoretically possible, it is strictly worse than eliminating \emph{mutual exclusion}, as it necessitates $t$ additional movements (discard-read-read versus read).
\begin{algorithm}[htb]
  \small
  \SetAlgoLined
  \KwGlobalIn{%
    $in[]$, 
    $out[]$, 
    $spine[]$, 
    $tileBump$, 
  }
  \KwSharedIn{%
    $wg\_partials[]$, 
    $wg\_fallback[]$,  
    $wg\_prev\_reduction$, 
    $wg\_broadcast$, 
    $wg\_control$, 
  }
  \KwGlobalOut{$out[]$ inclusive scan of $in[]$}
  \KwConstants{Vectors per thread $p$}

  \If{$threadid = 0$}{
    $wg\_broadcast \gets atomicAdd(\&tileBump, 1)$\;
    $wg\_control \gets LOCKED$\;
  }
  \textbf{barrier()}\;

  $part\_id \gets wg\_broadcast$\;

@@

  \textbf{Load}($in$, $part\_id$, \textit{local\_vectors})\;
  \textbf{SubgroupRake}(\textit{local\_vectors})\;

  \If{$highestRankLane(threadid)$}{
    $wg\_partials[subgroup\_id] \gets local\_vectors[final\_partial]$\;
  }
  \textbf{barrier()}\;

  \textbf{WorkgroupWideScan}($wg\_partials$)\;
  \textbf{barrier()}\;

  \tcp{Decoupled Fallback}
  \While{$wg\_control = LOCKED$}{
    \textbf{Lookback}($spine$, $wg\_prev\_reduction$, $wg\_control$)\;
    \textbf{barrier()}\;

    \If{$wg\_control = LOCKED$}{
      \textbf{FallbackReduce}($in$, $wg\_fallback$)\;
      \textbf{FallbackInsertionAttempt}($spine$, $wg\_fallback$, $wg\_prev\_reduction$, $wg\_control$)\;
      \textbf{barrier()}\;
    }
  }

  \textbf{Write}($out$, $part\_id$, \textit{local\_vectors}, $wg\_partials$, $wg\_prev\_reduction$)\;

  \caption{High-Level Scan Kernel}
\end{algorithm}
